---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ReciprocalGraphicalModels

<!-- badges: start -->
<!-- badges: end -->

## Intended use of the package

For a given dataset of gene expressions and DNA expressions, this R-package will estimate the gene-gene interactions and the gene-DNA interactions and then create a graph connecting them. A spike and slab prior is assumed for each interaction parameter based on our prior knowledge to calculate posterior estimates of the interactions. The package includes various helper functions that generate gibbs samples of the parameter values for the prior distributions and interaction parameter values based on other parameter values. The package have two main functions that, using the gibbs estimates of these parameters produced by the helper functions at each iteration, will provide estimates of the values of gene-gene interactions and gene-DNA interactions.

## Installation instructions

You can install ReciprocalGraphicalModels R package from GitHub with:

```{r, eval = F}

    install.packages("devtools")

    devtools::install_github("bitansa/ReciprocalGraphicalModels")

```  
      

Once the ReciprocalGraphicalModels library is installed load the library in the R workspace.

```{r}

    library("ReciprocalGraphicalModels")

```

## Example

We will give a small example to show the functionality of the two functions RGM and RGM_cpp available in this package to calculate gene-gene interactions and gene-DNA interactions for a given gene expressions and DNA expressions dataset. RGM_cpp function is a bit faster as it is written in Rcpp, but the functionality of both the functions are same. First we simulate the dataset. 

```{r}

set.seed(500)

# Number of datapoints
n = 500

# Number of Genes and number of DNAs
p = 3
k = 3

# Initialize gene-gene interaction matrix
A = matrix(sample(c(-3, 3), p^2, replace = TRUE), p, p)

# Diagonal entries of A matrix will always be 0
diag(A) = 0

# Initialize gene-DNA interaction matrix
B = matrix(0, p, k)

for(i in 1:p){

  B[i, i] = sample(c(-3, 3), 1, replace = TRUE)

}

# Indicator matrix for gene-DNA interaction
D = diag(3)

Sigma = 0.5 * diag(p)

Mult_Mat = solve(diag(p) - A)

Variance = Mult_Mat %*% Sigma %*% t(Mult_Mat)

# Generate DNA expressions
X = matrix(runif(n * k, 0, 5), nrow = n, ncol = k)

Y = matrix(0, nrow = n, ncol = p)

# Generate Gene expressions data based on DNA data
for (i in 1:n) {

 Y[i, ] = MASS::mvrnorm(n = 1, Mult_Mat %*% B %*% X[i, ], Variance)

}


```

Now we will apply RGM and RGM_cpp to show their functionality
```{r}

Output1 = RGM(X = X, Y = Y, A0 = A, B0 = B, D = D)

Output2 = RGM_cpp(X = X, Y = Y, A0 = A, B0 = B, D = D)

```

From the outputs we can get the gene-gene interactions in the following way:
```{r}

Output1$A

Output2$A

```

From the outputs we can get the gene-DNA interactions in the following way:
```{r}

Output1$B

Output2$B

```

We can also plot the log-likelihood to check the convergence of the model in the following way:
```{r}

plot(1:length(Output1$LL), Output1$LL, xlab = "Iterations", ylab = "Log-likelihood", type = 'l', main = "Loglikelihood plot VS Iterations")

plot(1:length(Output2$LL), Output2$LL, xlab = "Iterations", ylab = "Log-likelihood", type = 'l', main = "Loglikelihood plot VS Iterations")

```

Here in both the functions we have taken the initial starting point for A and B matrices to be the true A and B and hence the convergence the really fast. If we don't give any starting point for A and B or start with some random A and B a lot of iterations may be needed to reach convergence.


## Refrences
Yang Ni. Yuan Ji. Peter Müller. "Reciprocal Graphical Models for Integrative Gene Regulatory Network Analysis." Bayesian Anal. 13 (4) 1095 - 1110, December 2018. https://doi.org/10.1214/17-BA1087
