---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RGM

<!-- badges: start -->
[![R-CMD-check](https://github.com/bitansa/ReciprocalGraphicalModels/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bitansa/ReciprocalGraphicalModels/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## Intended use of the package

The RGM R package addresses a critical challenge in Mendelian randomization (MR) studies by offering a comprehensive solution. While MR has been instrumental in establishing causal relationships between exposures and outcomes, it typically focuses on the effects of single exposures on specific outcomes. However, biological systems are often complex, with interdependent outcomes influenced by numerous factors. RGM introduces a network-based approach to MR, allowing for exploration of the broader causal landscape. By leveraging network structures, RGM enables the construction of causal graphs that capture interactions between response variables and between responses and instrument variables.

RGM supports two models: Model 1, where the relationship is described by Y=AY+E, and Model 2, where it is Y=AY+BX+E. For Model 1, RGM accommodates individual-level data as well as summary-level data. For Model 2, RGM supports individual-level data and two types of summary-level data. This adaptability allows researchers to work with various data availability scenarios, enhancing the package's versatility.

The outputs of RGM include estimates of causal effects, adjacency matrices, and other valuable parameters. These outputs collectively contribute to a more comprehensive understanding of the intricate relationships within complex biological networks, thereby enhancing the insights gained from MR studies.




## Installation instructions

You can install RGM R package from GitHub with:

```{r, eval = F}

    install.packages("devtools")

    devtools::install_github("bitansa/RGM")

```  
      

Once the RGM package is installed load the library in the R workspace.

```{r}

     library("RGM")

```

## Example


We provide a concise illustration of the RGM function's capabilities within the package by demonstrating its use in calculating causal interactions between response variables and between response and instrument variables using simulated datasets. To initiate the example, we give two examples with Model 1, where Y=AY+E. In these examples, we generate synthetic datasets that mimic response variable expressions.

Subsequently, we offer three examples with Model 2, where Y=AY+BX+E. Here, we generate synthetic datasets that mimic response variable expressions alongside instrument variable expressions. These examples showcase the versatility of the RGM function, allowing researchers to explore and infer relationships between different response and instrument variables. Ultimately, this contributes to a deeper understanding of the underlying biological interactions.

```{r}

# Model 1: Y = AY + E

# Set seed
set.seed(9154)

# Number of datapoints
n = 10000

# Number of response variables 
p = 5

# Initialize causal interaction matrix between response variables
A = matrix(sample(c(-0.1, 0.1), p^2, replace = TRUE), p, p)

# Diagonal entries of A matrix will always be 0
diag(A) = 0

# Make the network sparse
A[sample(which(A!=0), length(which(A!=0))/2)] = 0


# Create variance-covariance matrix
Sigma = 1 * diag(p)

Mult_Mat = solve(diag(p) - A)

Variance = Mult_Mat %*% Sigma %*% t(Mult_Mat)

# Initialize response data matrix
Y = matrix(0, nrow = n, ncol = p)

# Generate response data matrix 
for (i in 1:n) {

 Y[i, ] = MASS::mvrnorm(n = 1, rep(0, p), Variance)

}

# Print true causal interaction matrices between response variables 
A


```

We will now apply RGM based on individual level data and summary level data to show its functionality.

```{r}

# Apply RGM on individual level data with Threshold prior for Model 1
Output1 = RGM(Model = "Model 1", Y = Y, n = 10000, prior = "Threshold")

# Calculate summary level data
S_YY = t(Y) %*% Y / n

# Apply RGM on summary level data for Spike and Slab Prior for Model 1
Output2 = RGM(Model = "Model 1", S_YY = S_YY,
              n = 10000, prior = "Spike and Slab")


```


We get the estimated causal interaction matrix between response variables in the following way:

```{r}

Output1$A_Est
Output2$A_Est


```
We get the estimated graph structure between the response variables in the following way:

```{r}

Output1$zA_Est
Output2$zA_Est


```

We can plot the log-likelihoods from the outputs in the following way:

```{r}

plot(Output1$LL_Pst, type = 'l', xlab = "Iterations", ylab = "Log-likelihood", col = 'red')
plot(Output2$LL_Pst, type = 'l', xlab = "Iterations", ylab = "Log-likelihood", col = 'blue')



```

Now, we will demonstrate RGM's functionality for Model 2.

```{r}

# Model 2: Y = AY + BX + E

# Set seed
set.seed(9154)

# Number of datapoints
n = 10000

# Number of response variables and number of instrument variables
p = 5
k = 6

# Create d vector
d = c(2, 1, 1, 1, 1)

# Initialize causal interaction matrix between response variables
A = matrix(sample(c(-0.1, 0.1), p^2, replace = TRUE), p, p)

# Diagonal entries of A matrix will always be 0
diag(A) = 0

# Make the network sparse
A[sample(which(A!=0), length(which(A!=0))/2)] = 0

# Initialize causal interaction matrix between response and instrument variables
B = matrix(0, p, k)

# Initialize m
m = 1

# Calculate B matrix based on d vector
for (i in 1:p) {
  
  # Update ith row of B
  B[i, m:(m + d[i] - 1)] = 1
  
  # Update m
  m = m + d[i]
  
}

# Create variance-covariance matrix
Sigma = 1 * diag(p)

Mult_Mat = solve(diag(p) - A)

Variance = Mult_Mat %*% Sigma %*% t(Mult_Mat)

# Generate instrument data matrix
X = matrix(runif(n * k, 0, 5), nrow = n, ncol = k)

# Initialize response data matrix
Y = matrix(0, nrow = n, ncol = p)

# Generate response data matrix based on instrument data matrix
for (i in 1:n) {

 Y[i, ] = MASS::mvrnorm(n = 1, Mult_Mat %*% B %*% X[i, ], Variance)

}

# Print true causal interaction matrices between response variables and between response and instrument variables
A
B


```

We will now apply RGM based on individual level data, summary level data and Beta, Sigma_Hat matrices to show its functionality.

```{r}

# Apply RGM on individual level data with Threshold prior
Output3 = RGM(Model = "Model 2", X = X, Y = Y, d = c(2, 1, 1, 1, 1), n = 10000, prior = "Threshold")

# Calculate summary level data
S_YY = t(Y) %*% Y / n
S_YX = t(Y) %*% X / n
S_XX = t(X) %*% X / n

# Apply RGM on summary level data for Spike and Slab Prior
Output4 = RGM(Model = "Model 2", S_YY = S_YY, S_YX = S_YX, S_XX = S_XX,
           d = c(2, 1, 1, 1, 1), n = 10000, prior = "Spike and Slab")

# Calculate Beta and Sigma_Hat
# Centralize Data
Y = t(t(Y) - colMeans(Y))
X = t(t(X) - colMeans(X))

# Calculate S_XX
S_XX = t(X) %*% X / n

# Generate Beta matrix and Sigma_Hat
Beta = matrix(0, nrow = p, ncol = k)
Sigma_Hat = matrix(0, nrow = p, ncol = k)

for (i in 1:p) {

    for (j in 1:k) {

        fit = lm(Y[, i] ~ X[, j])

        Beta[i, j] =  fit$coefficients[2]

        Sigma_Hat[i, j] = sum(fit$residuals^2) / n

        }

 }


# Apply RGM on S_XX, Beta and Sigma_Hat for Threshold Prior
Output5 = RGM(Model = "Model 2", S_XX = S_XX, Beta = Beta, Sigma_Hat = Sigma_Hat,
           d = c(2, 1, 1, 1, 1), n = 10000, prior = "Threshold")

```


We get the estimated causal interaction matrix between response variables in the following way:

```{r}

Output3$A_Est
Output4$A_Est
Output5$A_Est


```
We get the estimated graph structure between the response variables in the following way:

```{r}

Output3$zA_Est
Output4$zA_Est
Output5$zA_Est


```


We get the estimated causal interaction matrix between the response and the instrument variables from the outputs in the following way:

```{r}

Output3$B_Est
Output4$B_Est
Output5$B_Est


```
We get the estimated graph structure between the response and the instrument variables from the outputs in the following way:

```{r}

Output3$zB_Est
Output4$zB_Est
Output5$zB_Est


```

We can plot the log-likelihoods from the outputs in the following way:

```{r}

plot(Output3$LL_Pst, type = 'l', xlab = "Iterations", ylab = "Log-likelihood", col = 'orange')
plot(Output4$LL_Pst, type = 'l', xlab = "Iterations", ylab = "Log-likelihood", col = 'purple')
plot(Output5$LL_Pst, type = 'l', xlab = "Iterations", ylab = "Log-likelihood", col = 'cyan')


```

## References
Yang Ni. Yuan Ji. Peter Müller. "Reciprocal Graphical Models for Integrative Gene Regulatory Network Analysis." Bayesian Anal. 13 (4) 1095 - 1110, December 2018. https://doi.org/10.1214/17-BA1087
