---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RGM

<!-- badges: start -->
<!-- badges: end -->

## Intended use of the package

For a given dataset of gene expressions and DNA expressions, this R-package will estimate the gene-gene interactions and the gene-DNA interactions and then create a graph connecting them. A spike and slab prior is assumed for each interaction parameter based on our prior knowledge to calculate posterior estimates of the interactions. The package includes various helper functions that generate gibbs samples of the parameter values for the prior distributions and interaction parameter values based on other parameter values. The package have two main functions that, using the gibbs estimates of these parameters produced by the helper functions at each iteration, will provide estimates of the values of gene-gene interactions and gene-DNA interactions. 


## Installation instructions

You can install RGM R package from GitHub with:

```{r, eval = F}

    install.packages("devtools")

    devtools::install_github("bitansa/RGM")

```  
      

Once the RGM package is installed load the library in the R workspace.

```{r}

     library("RGM")

```

## Example

We will give a small example to show the functionality of the two functions RGM and RGM_cpp available in this package to calculate gene-gene interactions and gene-DNA interactions for a given gene expressions and DNA expressions dataset. RGM_cpp function is a bit faster as it is written in Rcpp, but the functionality of both the functions are same. First we simulate the dataset. 

```{r}

set.seed(9154)

# Number of datapoints
n = 10000

# Number of Genes and number of DNAs
p = 5
k = 6

# Indicator of which DNA affacts which gene
d = c(2, 1, 1, 1, 1)

# Initialize Protein-Protein interaction matrix
A = matrix(sample(c(-0.1, 0.1), p^2, replace = TRUE), p, p)

diag(A) = 0

# Make the network sparse
A[sample(which(A!=0), length(which(A!=0))/2)] = 0

# Initialize Protein-DNA interaction matrix
B = matrix(0, p, k)

# Initialize m
m = 1

# Calculate B matrix based on d vector
for (i in 1:p) {
  
  # Update ith row of B
  B[i, m:(m + d[i] - 1)] = 1
  
  # Update m
  m = m + d[i]
  
}

# Create variance-covariance matrix
Sigma = 1 * diag(p)

Mult_Mat = solve(diag(p) - A)

Variance = Mult_Mat %*% Sigma %*% t(Mult_Mat)

# Generate DNA expressions
X = matrix(runif(n * k, 0, 5), nrow = n, ncol = k)

Y = matrix(0, nrow = n, ncol = p)

# Generate Protein expressions data based on DNA data
for (i in 1:n) {

 Y[i, ] = MASS::mvrnorm(n = 1, Mult_Mat %*% B %*% X[i, ], Variance)

}

# Print Protein-Protein interactions and Protein-DNA interactions
A
B


```

We will now apply RGM based on individual level data, summary level data and beta matrix to show its functionality.

```{r}

# Apply RGM on individual level data with Threshold prior
Output1 = RGM(X = X, Y = Y, d = c(2, 1, 1, 1, 1), n = 10000, prior = "Threshold")

# Calculate summary level data
S_YY = t(Y) %*% Y / n
S_YX = t(Y) %*% X / n
S_XX = t(X) %*% X / n

# Apply RGM on summary level data for Spike and Slab Prior
Output2 = RGM(S_YY = S_YY, S_YX = S_YX, S_XX = S_XX,
           d = c(2, 1, 1, 1, 1), n = 10000, prior = "Spike and Slab")

# Calculate Beta and Sigma_Hat
# Centralize Data
Y = t(t(Y) - colMeans(Y))
X = t(t(X) - colMeans(X))

# Calculate S_XX
S_XX = t(X) %*% X / n

# Generate Beta matrix and Sigma_Hat
Beta = matrix(0, nrow = p, ncol = k)
Sigma_Hat = matrix(0, nrow = p, ncol = k)

for (i in 1:p) {

    for (j in 1:k) {

        fit = lm(Y[, i] ~ X[, j])

        Beta[i, j] =  fit$coefficients[2]

        Sigma_Hat[i, j] = sum(fit$residuals^2) / n

        }

}


# Apply RGM on S_XX, Beta and Sigma_Hat for Threshold Prior
Output3 = RGM(S_XX = S_XX, Beta = Beta, Sigma_Hat = Sigma_Hat,
           d = c(2, 1, 1, 1, 1), n = 10000, prior = "Threshold")

```

We get the Protein-Protein interactions from the outputs in the following way:
```{r}

Output1$A_Est
Output2$A_Est
Output3$A_Est


```
We get the graph structure between the proteins from the outputs in the following way:
```{r}

Output1$zA_Est
Output2$zA_Est
Output3$zA_Est


```

We get the Protein-DNA interactions from the outputs in the following way:
```{r}

Output1$B_Est
Output2$B_Est
Output3$B_Est


```
We get the graph structure between the proteins and the DNAs from the outputs in the following way:
```{r}

Output1$zB_Est
Output2$zB_Est
Output3$zB_Est


```

We can plot the log-likelihoods from the outputs in the following way:
```{r}

plot(Output1$LL_Pst, type = 'l', xlab = "Iterations", ylab = "Log-likelihood", col = 'red')
plot(Output2$LL_Pst, type = 'l', xlab = "Iterations", ylab = "Log-likelihood", col = 'blue')
plot(Output3$LL_Pst, type = 'l', xlab = "Iterations", ylab = "Log-likelihood", col = 'green')



```

## References
Yang Ni. Yuan Ji. Peter Müller. "Reciprocal Graphical Models for Integrative Gene Regulatory Network Analysis." Bayesian Anal. 13 (4) 1095 - 1110, December 2018. https://doi.org/10.1214/17-BA1087
